On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-21 manuscripts/PNAS/chazthm_pnas.tex
back to 6 pages

We  now show that $\pi$ is injective (in particular, a permutation if $\bar m = m$). Suppose that $\pi(i) = \pi(j) = \ell$ for some $i \neq j$ and $\ell$. Then, $\|c_{j}\mathbf{A}_{j} - \bar{c}_{j}\mathbf{B}_{\ell}\|_2 \leq \varepsilon$ and $\|c_{i}\mathbf{A}_{i} - \bar{c}_{i} \mathbf{B}_{\ell}\|_2  \leq \varepsilon$. Scaling and summing these inequalities by $|\bar{c}_{i}|$ and $|\bar{c}_{j}|$, respectively, and applying the triangle inequality, we obtain:
\begin{align*}
(|\bar{c}_{i}| + |\bar{c}_{j}|) \varepsilon
&\geq\|\mathbf{A}(\bar{c}_{i}c_{j} \mathbf{e}_{j} - \bar{c}_{j}c_{i}\mathbf{e}_{i})\|_2 \nonumber \\ 
&\geq  \left( |\bar{c}_{i}| + |\bar{c}_{j}| \right) L_2(\mathbf{A}) \min_{\ell \in [m]} |c_\ell |,
\end{align*}
which contradicts the bound $\varepsilon < L_2(\mathbf{A})/C_1$. Hence, the map $\pi$ is injective and therefore $\bar m \geq m$. Setting $\bar J = \pi([m])$ and letting $P = \left( \mathbf{e}_{\pi(1)} \cdots \mathbf{e}_{\pi(m)}\right)$ and $D = \text{diag}(\frac{\bar{c}_1}{c_1},\ldots,\frac{\bar{c}_m}{c_m})$, we see that \eqref{1D} becomes, for all $j \in [m]$:
\begin{align*}
\|\mathbf{A}_j - \mathbf{B}_{\bar J}\mathbf{PD}_j\|_2 
= \|\mathbf{A}_j - \frac{\bar{c}_j}{c_j}\mathbf{B}_{\pi(j)}\|_2 
\leq \frac{\varepsilon}{|c_j|} 
\leq C_1\varepsilon.
\end{align*}
\vspace{-.2 cm}
\end{proof}

@@

An immediate application of our theoretical work is Prop.~\ref{neuroprop}, which certifies the validity of perhaps the only published theory of neurally-plausible bottleneck communication in the brain: that sparse linear coding in a compressed space of neural activity can recover sparse codes sent through a randomly-constructed (but unknown) noisy wiring bottleneck \cite{Isely10}.\footnote{We refer the reader to \cite{ganguli2012compressed} for a discussion of this application's significance and more on the interactions between sparse linear coding and neuroscience.}

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). Assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data; moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, it seems a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis is the assumed well-posedness of parameter identification in the sparse linear coding model, confirmation of which forms the core of our theoretical findings.

@@

@@

\begin{lemma}\label{spanIntersectionLemma}
If $f: U \to V$ is an injective function then $f\left(\cap_{i=1}^\ell U_i \right) =  \cap_{i=1}^\ell f\left(U_i\right)$ for any $U_1, \ldots, U_\ell \subseteq U$.
%In particular, if for some $E \in 2^{[m]}$ the map $M \in \mathbb{R}^{n \times m}$ is injective on $\cup_{S \in \mathcal{H}} \text{span}\{e_i\}_{i \in S}$ then $\text{span}\{ M_{\cap_{S \in \mathcal{H}} S} \} = \cap_{S \in \mathcal{H}} \text{span}\{M_S\}$.
\end{lemma}
\begin{proof}
By induction it is enough to prove the case $|n| = 2$, but this case follows directly from the assumption. %Clearly, for any map $f$, if $w \in f(U \cap V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w \in f(U) \cap f(V)$. If $w \in f(U) \cap f(V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w = f(u) = f(v)$ for some $u \in U$ and $v \in V$, implying $u = v$ by injectivity of $f$. Hence $u \in U \cap V$, and $w \in f(U \cap V)$.
\end{proof}

\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $V_1, \ldots, V_k$ be subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (where $\xi$ is given in Def.~\ref{SpecialSupportSet}):
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{\xi(\{V_i\}_{i = 1}^k)} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).

@@

Finally, substituting this into \eqref{f} and rearranging produces \eqref{DTILeq} after replacing $1 - z$ with $\xi(\{V_i\}_{i=1}^k)$.
\end{proof}

\begin{lemma}\label{NonEmptyLemma} 
Fix an $r$-regular hypergraph $\mathcal{H} \subseteq 2^{[m]}$ satisfying the SIP. If the map $\pi: \mathcal{H} \to 2^{[\bar m]}$ has $\sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S|$ and:
\begin{align}\label{cond}
|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} |,\ \ \   \text{for } \mathcal{G} \subseteq \mathcal{H},
\end{align}
%
then $\bar m \geq m$; and if $\bar m < mr / (r-1)$, the association $i \mapsto \cap \pi(\sigma(i))$ is an injective map to $[\bar m]$ from some $J \subseteq [m]$ of size $\bar m - r(\bar m - m)$. %In particular, if $\bar m = m$ then $\pi$ is induced by a permutation on $[m]$.
\end{lemma}

>>>>

We  now show that $\pi$ is injective (in particular, a permutation if $\bar m = m$). Suppose that $\pi(i) = \pi(j) = \ell$ for some $i \neq j$ and $\ell$. Then, $\|c_{j}\mathbf{A}_{j} - \bar{c}_{j}\mathbf{B}_{\ell}\|_2 \leq \varepsilon$ and $\|c_{i}\mathbf{A}_{i} - \bar{c}_{i} \mathbf{B}_{\ell}\|_2  \leq \varepsilon$. Scaling and summing these inequalities by $|\bar{c}_{i}|$ and $|\bar{c}_{j}|$, respectively, and applying the triangle inequality, we obtain:
\begin{align*}
(|\bar{c}_{i}| + |\bar{c}_{j}|) \varepsilon
&\geq\|\mathbf{A}(\bar{c}_{i}c_{j} \mathbf{e}_{j} - \bar{c}_{j}c_{i}\mathbf{e}_{i})\|_2 \nonumber \\ 
&\geq  \left( |\bar{c}_{i}| + |\bar{c}_{j}| \right) L_2(\mathbf{A}) \min_{\ell \in [m]} |c_\ell |,
\end{align*}
which contradicts the bound $\varepsilon < L_2(\mathbf{A})/C_1$. Hence, the map $\pi$ is injective and therefore $\bar m \geq m$. Setting $\bar J = \pi([m])$ and letting $P = \left( \mathbf{e}_{\pi(1)} \cdots \mathbf{e}_{\pi(m)}\right)$ and $D = \text{diag}(\frac{\bar{c}_1}{c_1},\ldots,\frac{\bar{c}_m}{c_m})$, we see that \eqref{1D} becomes, for all $j \in [m]$:
\begin{align*}
\|\mathbf{A}_j - \mathbf{B}_{\bar J}\mathbf{PD}_j\|_2 
= \|\mathbf{A}_j - \frac{\bar{c}_j}{c_j}\mathbf{B}_{\pi(j)}\|_2 
\leq \frac{\varepsilon}{|c_j|} 
\leq C_1\varepsilon.
\end{align*}
%\vspace{-.2 cm}
\end{proof}

@@

An immediate application of our theoretical work is Prop.~\ref{neuroprop}, which certifies the validity of perhaps the only published theory of neurally-plausible bottleneck communication in the brain: that sparse linear coding in a compressed space of neural activity can recover sparse codes sent through a randomly-constructed (but unknown) noisy wiring bottleneck \cite{Isely10}.\footnote{We refer the reader to \cite{ganguli2012compressed} for more on interactions between dictionary learning and neuroscience.}

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). Assurances such as those offered by our theorems certify that different devices (with different initialization, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, it seems a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis is the assumed well-posedness of parameter identification in the sparse linear coding model, confirmation of which forms the core of our theoretical findings.

@@

\vspace{-.15 cm}

@@

\begin{lemma}\label{spanIntersectionLemma}
If $f: U \to V$ is an injective function then $f\left(\cap_{i=1}^\ell U_i \right) =  \cap_{i=1}^\ell f\left(U_i\right)$ for any $U_1, \ldots, U_\ell \subseteq U$.
%In particular, if for some $E \in 2^{[m]}$ the map $M \in \mathbb{R}^{n \times m}$ is injective on $\cup_{S \in \mathcal{H}} \text{span}\{e_i\}_{i \in S}$ then $\text{span}\{ M_{\cap_{S \in \mathcal{H}} S} \} = \cap_{S \in \mathcal{H}} \text{span}\{M_S\}$.
\end{lemma}
\begin{proof}
By induction it is enough to prove the case $|n| = 2$, but this case follows directly from the assumption. %Clearly, for any map $f$, if $w \in f(U \cap V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w \in f(U) \cap f(V)$. If $w \in f(U) \cap f(V)$ then $w \in f(U)$ and $w \in f(V)$, hence $w = f(u) = f(v)$ for some $u \in U$ and $v \in V$, implying $u = v$ by injectivity of $f$. Hence $u \in U \cap V$, and $w \in f(U \cap V)$.
\end{proof}
%\vspace{-.2 cm}
\begin{lemma}\label{DistanceToIntersectionLemma}
Fix $k \geq 2$. Let $V_1, \ldots, V_k$ be subspaces of $\mathbb{R}^m$ and set $V = \cap_{i = 1}^k V_i$. For  $\mathbf{x} \in \mathbb{R}^m$, we have (where $\xi$ is given in Def.~\ref{SpecialSupportSet}):
\vspace{-.2 cm}
\begin{align}\label{DTILeq}
\text{\rm dist}(\mathbf{x}, V) \leq \frac{1}{\xi(\{V_i\}_{i = 1}^k)} \sum_{i=1}^k \text{\rm dist}(\mathbf{x}, V_i).
\end{align}
\end{lemma}
%\vspace{-.3 cm}
\begin{proof} 
Recall the orthogonal projection onto a subspace $V \subseteq \mathbb{R}^m$ is the mapping $\Pi_V: \mathbb{R}^m \to V$ that associates with each $\mathbf{x}$ its unique nearest point in $V$; i.e., $\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 = \text{\rm dist}(\mathbf{x}, V)$.
Next, observe:
\begin{align}\label{f}
\|\mathbf{x} - \Pi_V\mathbf{x}\|_2 &\leq \|\mathbf{x} - \Pi_{V_k} \mathbf{x}\|_2 + \|\Pi_{V_k}  \mathbf{x} - \Pi_{V_k}\Pi_{V_{k-1}}\mathbf{x}\|_2 \nonumber \\
&\ \ \ + \cdots + \|\Pi_{V_k} \Pi_{V_{k-1}}\cdots \Pi_{V_1} \mathbf{x} - \Pi_V \mathbf{x}\|_2 \nonumber \\
&\leq \|\Pi_{V_k}\cdots\Pi_{V_{1}} \mathbf{x} - \Pi_V \mathbf{x}\|_2 + \sum_{\ell=1}^k \|\mathbf{x} - \Pi_{V_{\ell}} \mathbf{x}\|_2,
\end{align}
using the triangle inequality and that the spectral norm satisfies $\|\Pi_{V_{\ell}}\|_2 \leq 1$ for all $\ell$ (since $\Pi_{V_{\ell}}$ are orthogonal projections).

@@

Finally, substituting this into \eqref{f} and rearranging produces \eqref{DTILeq} after replacing $1 - z$ with $\xi(\{V_i\}_{i=1}^k)$.
\end{proof}
%\vspace{-.4 cm}
\begin{lemma}\label{NonEmptyLemma} 
Fix an $r$-regular hypergraph $\mathcal{H} \subseteq 2^{[m]}$ satisfying the SIP. If the map $\pi: \mathcal{H} \to 2^{[\bar m]}$ has $\sum_{S \in \mathcal{H}} |\pi(S)| \geq \sum_{S \in \mathcal{H}} |S|$ and:
\begin{align}\label{cond}
|\cap \pi(\mathcal{G})| \leq |\cap \mathcal{G} |,\ \ \   \text{for } \mathcal{G} \subseteq \mathcal{H},
\end{align}
%
then $\bar m \geq m$; and if $\bar m < mr / (r-1)$, the association $i \mapsto \cap \pi(\sigma(i))$ is an injective map to $[\bar m]$ from some $J \subseteq [m]$ of size $\bar m - r(\bar m - m)$. %In particular, if $\bar m = m$ then $\pi$ is induced by a permutation on $[m]$.
\end{lemma}
